\documentclass[a4paper,12pt,oneside,ms]{memoir}

\usepackage{amssymb}
\usepackage[svgnames]{xcolor}

\usepackage{pf2}
\usepackage{abbrev}


\pagecolor{Lavender}
\pagestyle{empty}
\raggedright

\begin{abbreviate}{FOO}{|[]`.}{|[]`.}
  \x |¨|  {\bigcap }
  \x |¨   {|¨}
  \x |  {|}

  \x `'F {F^{-1}}
  \x `' {`'}
  \x ` {`}
  \x [ {\langle}
  \x ] {\rangle}
  \x .A {\forall}
  \x .e {\in}
  \x . {.}
\end{abbreviate}

\begin{document}

\centerline{\textsc{Inpreima2}}
\smallskip
\centerline{FL}
\smallskip
\centerline{Tuesday 5 February 2012}
% \medskip

\FOO
\begin{flushleft}\colorbox{PaleGoldenrod}{\begin{minipage}{\textwidth}
\noindent\textsc{Theorem:}
The preimage of the intersection of a non-empty set $A$ is 
the intersection of the preimage of the elements of $A$.
\end{minipage}}\end{flushleft} 

\begin{flushleft}
Remark: I write
$F(x)$ when $x$ is an element of the domain of $F$ and $F[x]$
when $x$ is a part of the domain of $F$.
\end{flushleft} 

\noFOO
% \medskip

\FOO
\centerline{\rule{5cm}{1pt}}
\bigskip
\assume{
  \begin{pfenum}
    \item $A \neq \varnothing$ \label{in1}
    \item $F$ is a function \label{in2}
  \end{pfenum}}
  \prove{$ `'F  [ |¨| A ] = |¨| _{x .e A} `'F [ x ]$}
\begin{proof}
  \pflet{y be a set}
  \step{<3>1}{\suffices{$y .e  `'F  [ |¨| A ]$ iff $y .e |¨| _{x  .e  A} `'F [ x ]$}}
  \begin{proof}
    By definition of equality (see eqriv).
  \end{proof}
  \step{<3>2}{$y .e |¨| _{x  .e  A}  `'F [ x ]$ iff  $.A x .e A\; y .e `'F [ x ]$}
  \begin{proof}
    By definition of an intersection (eliin).
  \end{proof}
  \step{<3>3}{$y  .e  `'F [  |¨| A ] $ iff $ .A  x  .e  A\; y  .e   `'F  [  x  ] $}
  \begin{proof}
    \step{<5>1}{If $y  .e   `'F  [ |¨|  A ] $ then $ .A  x  .e  A\; y  .e   `'F  [  x  ] $.}
    \begin{proof}
      \step{<6>1}{\assume{$y  .e   `'F  [ |¨|  A ] $}\prove{$ .A  x  .e  A\; y  .e   `'F  [  x  ] $}}
      \begin{proof}
        \step{<7>1}{$F(y)  .e  |¨|  A$}
        \begin{proof}
          By  fvimacnv and assumption \stepref{<6>1}.
        \end{proof}
        \step{<7>2}{$ .A  x  .e  A\; F(y)  .e  x$ by \stepref{<7>1} and
          definition of an intersection (elint).}
        \qedstep
        \begin{proof}
          By  fvimacnv and \stepref{<7>2}.
        \end{proof}
      \end{proof}
      \qedstep
      \begin{proof}
        By deduction theorem.
      \end{proof}
    \end{proof}
    \step{<5>2}{If $ .A  x  .e  A\; y  .e   `'F  [  x  ] $ then $y  .e   `'F  [ |¨|  A ] $.}
    \begin{proof}
      \step{<6>1}{\assume{$ .A  x  .e  A\; y  .e   `'F  [  x  ] $}
                  \prove{$y  .e   `'F  [ |¨|  A ] $}}
      \begin{proof}
        \step{<7>1}{$ .A  x  .e  A\; F(y)  .e  x$ (by assumption \stepref{<6>1} 
          and fvimacnvi.)}
        \step{<7>2}{$F(y)  .e  |¨|  A$ (by \stepref{<7>1} and definition
          of an intersection (elint2).)}
        \qedstep 
        \begin{proof}
           By \stepref{<7>2} and fvimacnv.
        \end{proof}
      \end{proof}
      \qedstep
      \begin{proof}
        By deduction theorem.
      \end{proof}
    \end{proof}
    \qedstep
    \begin{proof}
      By \stepref{<5>1} and \stepref{<5>2}.
    \end{proof}
  \end{proof}
  \qedstep
  \begin{proof}
    By  \stepref{<3>2} and \stepref{<3>3}.
  \end{proof}
\end{proof}

\noFOO
\centerline{\rule{5cm}{1pt}}
\medskip

\textsc{Theorem fvimacnvi:} \FOO If $F$ is a function and $A .e `'F [
B ]$ then $F(A) .e B$. \noFOO
\medskip

\textsc{Theorem fvimacnv:} \FOO If $F$ is a function and $A$ an element of its
domain then $A  .e  `'F [  B  ] $ iff $F(A)  .e  B$. \noFOO
\medskip
\newpage
\textsc{Theorem elint2:} \FOO If $A$ is a set, and $B$ a class, 
$A .e |¨| B$ iff $.A x .e B\; A .e x$. \noFOO
\medskip

\centerline{\rule{5cm}{1pt}}
\medskip

Simple propositional  calculus tautologies  may be used  without being
explicitely mentionned. The ``deduction''  theorem may be used without
being explicitely mentionned.

\bigskip
The keyword ``suffices'' changes the  goal. The goal is always mention
at the  upper level. When ``suffices''  is used this  goal changes and
the proof  behind the ``suffices'' step  ensures the new  goal and the
former one  are equivalent. The other  steps at the  same level behind
the ``suffices'' are used to prove  the new goal. The ``Q.E.D'' at the
end of the level refers to the new goal.

\bigskip
The system used is in the Gentzen style. Let's recall set.mm is in
the Hilbert style.

\centerline{\rule{5cm}{1pt}}
\medskip

Modified on 3 June 2013 and again 13 July 2016.

\end{document}
